On the Minimum Core Mass for Giant Planet Formation

### 2013-10-16 tex/draft/draft4.tex
section 3 minor edits

This section develops the analytic version of our two-layer model, which is useful for predicting and explaining the results of numerical evolution models.  To obtain our analytic results, we neglect self-gravity.   As shown in Section \ref{sec:KH}, self-gravity becomes quantitatively important at surprisingly low atmospheric masses, $M_{\rm atm} \gtrsim 0.2 M\co$.  Thus the analytic model cannot accurately model the the later stages of evolution.  Nevertheless, the analytic model does successfully explain general trends with opacity and disk temperature and pressure, for example.

The analytic model also assumes that the upper radiative radiative layer is thick, with a pressure depth $P\cb \gg P\di$.  This approximation limits the ability to study the initial cooling of very low mass atmospheres with thin outer radiative zones.  However since early evolution is quite rapid, this approximation negligibly affects total cooling times and criticial core masses, our main foci.

@@

If the radiative zone were replaced by an isothermal layer, then hydrostatic balance would give $\chi = \theta = 1$.  Standard radiative structure calculations (see \App{RCBcorr} for details) give 
\begin{equation}
\label{eq:chi}
\chi \simeq \Big(1-\frac{\delad}{\nabla_{\infty}} \Big)^{-\frac{1}{4-\beta}} \simeq 1.53 \, ,
\end{equation}
for $P\cb \gg P\di$, our regime of interest, and with the radiative temperature gradient at infinite depth, $\nabla_\infty = 1/2$, for our dust opacity.  The radiative zones are thus nearly isothermal, as also found by R06.  While  a simple closed form expression is not possible, the factor $\theta \simeq 0.556$ for our parameters.   According to \Eq{eq:PcbRcb} our pressure criterion also gives $R\cb \ll \RB$, but the inequality is logarithmically weaker.

Given the conditions at the RCB, hydrostatic balance gives the density and temperature profiles in the inner region along the adiabat,
\begin{subeqnarray}
\rho &=& \rho\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right]^{1/(\gamma -1)} \slabel{eq:rhoconv}  \\
T	&=& T\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right] \slabel{eq:Tconv}\, .
\end{subeqnarray} 
%Note (to self) that the "1+" is neglibible in Rcb << RB limit taken elsewhere. 
The effective Bondi radius
\begin{equation}
\RB' \equiv {G M\co \over C_P T\cb} = {\delad \over \chi} \RB
\end{equation} 
simplifies expressions, with $C_P = \Rg / \delad$ the specific heat at constant pressure.

Even simpler expressions hold deep in the adiabatic interior, where $r \ll R\cb \ll \RB'$ gives
\begin{subeqnarray}\label{eq:deep}
\rho &\simeq& \rho\cb \left[ {\RB' \over r}\right]^{1/(\gamma -1)} \propto r^{-5/2} \slabel{eq:rhoconvdeep}  \\
T	&\simeq& T\cb {\RB' \over r} = {G M\co \over C_P r} \slabel{eq:Tconvdeep}\, .
\end{subeqnarray} 
While the radial density profile depends on the adiabatic index, the temperature scaling does not.  For reference, if self-gravity is included, the temperature profile satifies
\begin{equation} \label{eq:TPsg}
{dT \over dr} = - {G m(r) \over C_P r^2}\, .
\end{equation} 
Thus the enclosed mass in self-gravitating models gives a temperature profile flatter than $T \propto r^{-1}$.

The total specific energy at depth
\begin{equation}\label{eq:ean}
e = e_g + u = -\delad {GM\co \over r} \, .
\end{equation} 
is thus generically proportional to the gravitational potential, $e_g = -GM\co/r$. 
%AY: note $u$ introduced already in section 2.  Furthermore meaning obvious from context here.  e_G is good to restate in the non-self-grav limit which changed.

\subsection{Mass, Energy and Luminosity}
\label{MELan}
The most relevant quantities for the global cooling model (equation \ref{eq:coolingglobal}) are the integrated energy, luminosity, and atmospheric mass.  In our non-self-gravitating limit, the mass of our nearly isothermal radiative zones is less than the convective interior, as shown in \App{iso}.

\Eq{eq:rhoconv} thus integrates to give the atmospheric mass
\begin{eqnarray} 
\label{eq:Matman}
M_{\rm atm} %&=& 4 \pi \int_{R\co}^{R\cb} \rho r^2 dr \\  %AY: any interested reader knows this step, right?
&=& {5 \pi^2 \over 4} \rho\cb {\RB'}^{5/2} \sqrt{R\cb}, 
%AY: alternate forms
% \approx 0.54 \rho\cb {\RB'}^{5/2} \sqrt{r\cb} \\
%&=& 0.54{\RB'^{3} \sqrt{\chi} \over  \sqrt{ \ln[P\cb/(\theta P\di)]}} \, .
\end{eqnarray}
in the relevant limit $R\co \ll R\cb \ll \RB'$ and for $\gamma =  7/5$.  Mass is concentrated near the outer regions of the convective zone, a result that holds more generally for $\gamma > 4/3$.

@@

The total energy is concentrated towards the core if $u \rho  r^3 \propto \rho r^2$ drops with increasing $r$.  This condition requires $\gamma < 3/2$, which our choice of $\gamma = 7/5$ satisfies, but $\gamma = 5/3$ does not. 
Integration of \Eq{eq:ean} over the mass of the atmosphere thus gives
\begin{subeqnarray} 
E &=& - 4 \pi \nabla_{\rm ad} G M\co \int_{R\co}^{R\cb} \rho r dr \\
&\approx& - 4 \pi P\cb {\RB'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \slabel{eq:Ean} \\ 
&\approx& - 8 \pi P\cb {\RB'^{7/2} \over \sqrt{R\co}} \slabel{eq:Eus}
%&\approx& -0.31 P\cb {(\RB / \chi)^{7/2} \over \sqrt{R\co}}
%&\approx& -0.31 P\cb {\RB'^{7/2} \over \sqrt{R\co}}
\end{subeqnarray} 
with $\gamma < 3/2$ and $\gamma = 7/5$ in \Eqs{eq:Ean}{eq:Eus}, respectively. % and then revert to the standard Bondi radius $\RB = G M\co/(\Rg T\di)$ in terms of the disk temperature.

Applying marginal convective stability, $\delrad = \delad$, to \Eq{eq:delrad} gives the emergent luminosity from the radiative-convective boundary, 
\begin{equation} \label{eq:Lcb}
L\cb = {64 \pi G M\cb \sigma T\cb^4 \over 3 \kappa P\cb } \nabla_{\rm ad} \approx L\di {P_{\rm d} \over P\cb}\, , % \left({P\di \over P\cb}\right)%^{1+\alpha}
\end{equation} 
where we define 
\begin{equation} 
L\di \equiv {64 \pi G M\cb \sigma T_{\rm d}^4 \over 3 \kappa(T_{\rm d}) P_{\rm d}} \nabla_{\rm ad}\chi^{4-\beta}.
\end{equation} 
The inverse scaling $L\cb \propto 1/P\cb$ shows that luminosity drops as the atmosphere cools and $P\cb$ deepens.  This result relies on the pressure independence of  dust opacities.  For fully non-self-gravitating results, we replace $M\cb$, the mass up to the RCB, with the core mass, but the mass of the convective atmosphere could be included for a slightly higher order estimate.

\subsection{Simplified Cooling Model}
\label{coolingan}
%We can eliminate $R\cb$ from equation (\ref{eq:Matman}) using \Eq{eq:PcbRcb}.  The ratio of atmosphere to core mass is then  
%\begin{equation} \label{eq:crit}
%{M_{\rm atm} \over M\co} \approx {P\cb /P_M \over  \sqrt{\ln[ P\cb/(\theta P_{\rm d})]}},
%\end{equation} 
%where we introduce a characteristic pressure
%\begin{equation} 
%P_M \equiv {4 \delad^{3/2} \over 5 \pi^2 \sqrt{\chi} } {G M\co^2 \over {\RB'}^4}\, .
%%P_M \approx {1.85 \over \sqrt{\chi}} {G M\co^2 \over {\RB'}^4}\, .
%\end{equation} 
%For $M_{\rm atm} = M\co$ in equation (\ref{eq:crit}) we require
%\begin{equation} \label{eq:Pcbc}
%P\cb = \xi P_M ,
%\end{equation} 
%where  the logarithmic factor
%\begin{equation}\label{eq:xi}
%\xi \equiv \sqrt{\ln[ P\cb/(\theta P_{\rm d})]} = \sqrt{\ln[ \xi P_M / (\theta P_{\rm d})]}
%\end{equation} 
%is found by numerically solving the above transcendental equation. The physical solution has $\xi >1$, but is typically of order unity. Numerical solutions exist for $\xi \ll 1$, but are unphysical as they imply $P\cb < P_{\rm d}$.

@@

Both $t_{\rm co}$ and $\MC$ are too large to be interesting, or to be correct based on previous results and the numerical results in this Paper.  The main reason for this discrepancy is the neglect of self-gravity, as shown in Section \ref{sec:KH}.  The magnitude of the discrepancy may be surprising as one might expect self-gravity to be a modest factor for $M_{\rm atm} \leq M\co$.  Clearly this seemingly reasonable expectation must be applied with caution, an interesting result in itself.

The analytic model is also useful because, despite the amplitude error, it explains the basic scaling of our (and other) numerical models.  As is well known, a lower opacity, here scaling with $F_\kappa$, accelerates growth and gives a lower critical core mass. Lower disk temperatures also decrease both $t_{\rm co}$ and $\MC$.  Furthermore the decline in both quantities with semimajor axis is completely generated by the disk temperature law (ignoring $\xi$).  The temperature dependence are a competition between two main effects.  The higher luminosity, $\propto T^{4-\beta}$, at higher temperatures gives faster cooling, which opposes the overall effect.  The larger Bondi radius, and the effectively stronger gravity of the core, is the dominant effect that gives faster cooling at lower temperatures.  For different $\beta$ values, $t_{\rm co} \propto F_T^{ \beta + 1/2}$, so the slope of the dust opacity is an important moderating effect on the cooling.

@@

\begin{equation}
\label{eq:delo}
\nabla \di \equiv \frac{3 \kappa(T{\di}) P{\di}}{64 \pi G M \sigma T_{\rm d}^4} L.
\end{equation}
Here $M$ is the total planet mass. Since our analytic model neglects self-gravity, $M=M\co$ and therefore $\nabla\di$ is constant. From equation (\ref{eq:delradan}) and $\delrad=d \ln T / d\ln P$, the temperature profile in the radiative region integrates to
\begin{equation}\label{eq:radTP}
\left(T \over T_{\rm d}\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left(\left{P \over P_{\rm d}}\right - 1 \right) \, ,
\end{equation} 
where $\nabla_\infty = 1/(4-\beta)$ is the radiative temperature gradient for $T ,P \rightarrow \infty$.
Applying \Eqs{eq:delradan}{eq:radTP} at the RCB (where $\delrad = \delad$) under the assumption that $P\cb \gg P_{\rm d}$ results in  $T\cb=\chi T\di$ as in \Eq{eq:Tcb}, with $\chi$ defined in \Eq{eq:chi}.

@@

\subsection{Surface Terms}
\label{surfterms}
In this section we check the relevance of the neglected surface terms in \Eq{eq:coolingglobal}.  We first show that accretion energy is only a small correction at the RCB, which is where we apply our cooling model. A rough comparison (ignoring terms of order $\xi$) of  accretion luminosity vs. $\dot{E}$ gives
\begin{equation}
{G M \dot{M} \over R \dot{E}} = {G M  \over R}{dM \over dE} \sim {G M\co^2 \over R_B E}{P\cb \over P_M} \sim \sqrt{R\co \over R_B} \ll 1,
\end{equation} 
where we assume $P\cb \sim P_M$  for a massive atmosphere.  Accretion energy at the protoplanetary surface is thus very weak for marginally self-gravitating atmospheres, and even weaker for lower mass atmospheres.  A similar scaling analysis shows that the work term $P\surf \p V\surf/\p t$ is similarly weak.  Nevertheless, our numerical calculations include these surface terms in a more realistic and complete model of self-gravitating atmospheres.

>>>>

This section develops the analytic version of our two-layer model, which is useful for predicting and explaining the results of numerical evolution models.  To obtain our analytic results, we neglect self-gravity.   As shown in Section \ref{sec:KH}, self-gravity becomes quantitatively important at surprisingly low atmospheric masses, $M_{\rm atm} \gtrsim 0.2 M\co$.  Thus the analytic model cannot accurately model the later stages of evolution.  Nevertheless, this model does explain general trends, including the dependence on opacity and disk temperature and pressure.

The analytic model also assumes that the upper radiative radiative layer is thick, with a pressure depth $P\cb \gg P\di$.  This approximation limits the ability to study the initial cooling of very low mass atmospheres with thin outer radiative zones.  However since early evolution is quite rapid, this approximation negligibly affects total cooling times and critical core masses, our main foci.

@@

The leading constants would be unity, $\chi = \theta = 1$, if the radiative zone were replaced by an isothermal layer.  In practice deviations from unity are modest.    Standard radiative structure calculations (see \App{RCBcorr} for details) give 
\begin{equation}
\label{eq:chi}
\chi \simeq \Big(1-\frac{\delad}{\nabla_{\infty}} \Big)^{-\frac{1}{4-\beta}} \simeq 1.53 \, ,
\end{equation}
for $P\cb \gg P\di$, our regime of interest, and with the radiative temperature gradient at infinite depth, $\nabla_\infty = 1/2$, for our dust opacity.  The radiative zones are thus nearly isothermal, as also found by R06.  While  a simple closed form expression is not possible, the factor $\theta \simeq 0.556$ for our parameters.   
%According to \Eq{eq:PcbRcb} our pressure criterion also gives $R\cb \ll \RB$, but the inequality is logarithmically weaker.

Given the conditions at the RCB,  the density and temperature profiles along the interior adiabat,
\begin{subeqnarray}
\rho &=& \rho\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right]^{1/(\gamma -1)} \slabel{eq:rhoconv}  \\
T	&=& T\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right] \slabel{eq:Tconv}\, ,
\end{subeqnarray} 
%Note (to self) that the "1+" is neglibible in Rcb << RB limit taken elsewhere. 
follow from hydrostatic balance.  We introduce an effective Bondi radius
\begin{equation}
\RB' \equiv {G M\co \over C_P T\cb} = {\delad \over \chi} \RB
\end{equation} 
to simplify expressions, with $C_P = \Rg / \delad$ the specific heat at constant pressure.

Deep in the adiabatic interior, where $r \ll R\cb \ll \RB'$, the profiles follow simple power laws,
\begin{subeqnarray}\label{eq:deep}
\rho &\simeq& \rho\cb \left[ {\RB' \over r}\right]^{1/(\gamma -1)} \propto r^{-5/2} \slabel{eq:rhoconvdeep}  \\
T	&\simeq& T\cb {\RB' \over r} = {G M\co \over C_P r} \slabel{eq:Tconvdeep}\, .
\end{subeqnarray} 
While the radial density profile depends on the adiabatic index, the $r^{-1}$ temperature scaling is universal.  For reference, if self-gravity is included, the temperature profile satisfies
\begin{equation} \label{eq:TPsg}
{dT \over dr} = - {G m(r) \over C_P r^2}\, .
\end{equation} 
Thus in self-gravitating models, the enclosed mass gives a temperature profile flatter than $T \propto r^{-1}$.

Returning to our non-self-gravitating model, the total specific energy at depth
\begin{equation}\label{eq:ean}
e = e_g + u = -\delad {GM\co \over r} \, .
\end{equation} 
is simply proportional to the gravitational potential, $e_g = -GM\co/r$.

\subsection{Mass, Energy and Luminosity}
\label{MELan}
The most relevant quantities for global cooling are the integrated energy, luminosity, and atmospheric mass.  In our non-self-gravitating limit, the mass of our nearly isothermal radiative zones is less than the convective interior, as shown in \App{iso}.

The atmospheric mass is thus given by the integration of \Eq{eq:rhoconv}, 
\begin{eqnarray} 
\label{eq:Matman}
M_{\rm atm} %&=& 4 \pi \int_{R\co}^{R\cb} \rho r^2 dr \\  %AY: any interested reader knows this step, right?
&=& {5 \pi^2 \over 4} \rho\cb {\RB'}^{5/2} \sqrt{R\cb}, 
%AY: alternate forms
% \approx 0.54 \rho\cb {\RB'}^{5/2} \sqrt{r\cb} \\
%&=& 0.54{\RB'^{3} \sqrt{\chi} \over  \sqrt{ \ln[P\cb/(\theta P\di)]}} \, .
\end{eqnarray}
in the relevant limit $R\co \ll R\cb \ll \RB'$ and for $\gamma =  7/5$.  Mass is concentrated near the outer regions of the convective zone, a result that holds for $\gamma > 4/3$.

@@

The total energy is concentrated towards the core if $|u| \rho  r^3 \propto \rho r^2$ drops with increasing $r$.  This condition requires $\gamma < 3/2$, which our choice of $\gamma = 7/5$ satisfies, but $\gamma = 5/3$ would not. 
Integration of \Eq{eq:ean} over the mass of the atmosphere thus gives
\begin{subeqnarray} 
E &=& - 4 \pi \nabla_{\rm ad} G M\co \int_{R\co}^{R\cb} \rho r dr \\
&\approx& - 4 \pi P\cb {\RB'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \slabel{eq:Ean} \\ 
&\approx& - 8 \pi P\cb {\RB'^{7/2} \over \sqrt{R\co}} \slabel{eq:Eus}
%&\approx& -0.31 P\cb {(\RB / \chi)^{7/2} \over \sqrt{R\co}}
%&\approx& -0.31 P\cb {\RB'^{7/2} \over \sqrt{R\co}}
\end{subeqnarray} 
with $\gamma < 3/2$ and $\gamma = 7/5$ in \Eqs{eq:Ean}{eq:Eus}, respectively. % and then revert to the standard Bondi radius $\RB = G M\co/(\Rg T\di)$ in terms of the disk temperature.

The emergent luminosity from the radiative-convective boundary, 
\begin{equation} \label{eq:Lcb}
L\cb = {64 \pi G M\cb \sigma T\cb^4 \over 3 \kappa P\cb } \nabla_{\rm ad} \approx L\di {P_{\rm d} \over P\cb}\, , % \left({P\di \over P\cb}\right)%^{1+\alpha}
\end{equation} 
follows from \Eq{eq:delrad} and marginal convective stability, $\delrad = \delad$, where we define 
\begin{equation} 
L\di \equiv {64 \pi G M\cb \sigma T_{\rm d}^4 \over 3 \kappa(T_{\rm d}) P_{\rm d}} \nabla_{\rm ad}\chi^{4-\beta}.
\end{equation} 
The inverse scaling $L\cb \propto 1/P\cb$ shows that luminosity drops as the atmosphere cools and $P\cb$ deepens.  This result relies on the pressure independence of  dust opacities.  For fully non-self-gravitating results, we replace $M\cb$, the mass up to the RCB, with the core mass, but the mass of the convective atmosphere can be included for a slightly higher order estimate.

\subsection{Cooling Times \& Core Masses}
\label{coolingan}
%We can eliminate $R\cb$ from equation (\ref{eq:Matman}) using \Eq{eq:PcbRcb}.  The ratio of atmosphere to core mass is then  
%\begin{equation} \label{eq:crit}
%{M_{\rm atm} \over M\co} \approx {P\cb /P_M \over  \sqrt{\ln[ P\cb/(\theta P_{\rm d})]}},
%\end{equation} 
%where we introduce a characteristic pressure
%\begin{equation} 
%P_M \equiv {4 \delad^{3/2} \over 5 \pi^2 \sqrt{\chi} } {G M\co^2 \over {\RB'}^4}\, .
%%P_M \approx {1.85 \over \sqrt{\chi}} {G M\co^2 \over {\RB'}^4}\, .
%\end{equation} 
%For $M_{\rm atm} = M\co$ in equation (\ref{eq:crit}) we require
%\begin{equation} \label{eq:Pcbc}
%P\cb = \xi P_M ,
%\end{equation} 
%where  the logarithmic factor
%\begin{equation}\label{eq:xi}
%\xi \equiv \sqrt{\ln[ P\cb/(\theta P_{\rm d})]} = \sqrt{\ln[ \xi P_M / (\theta P_{\rm d})]}
%\end{equation} 
%is found by numerically solving the above transcendental equation. The physical solution has $\xi >1$, but is typically of order unity. Numerical solutions exist for $\xi \ll 1$, but are unphysical as they imply $P\cb < P_{\rm d}$.

@@

Both $t_{\rm co}$ and $\MC$ are too large to be interesting, or to be correct based on previous results and the numerical results in this Paper.  The main reason for this discrepancy is the neglect of self-gravity, as shown in Section \ref{sec:KH}.  The magnitude of the discrepancy may be surprising as one might expect self-gravity to be a modest factor for $M_{\rm atm} \leq M\co$.  Clearly this seemingly reasonable expectation can be misleading, an interesting result in itself.

Aside from this insight, the analytic model is useful because, despite the amplitude error, it explains the basic scaling of numerical cooling models.  As is well known, a lower opacity, here scaling with $F_\kappa$, accelerates growth and gives a lower critical core mass. Lower disk temperatures also decrease both $t_{\rm co}$ and $\MC$.  Furthermore the decline in both quantities with semimajor axis is completely generated by the disk temperature law (ignoring the logarithmic $\xi$).  The temperature dependence are a competition between two main effects.  The higher luminosity, $\propto T^{4-\beta}$, at higher temperatures gives faster cooling, which opposes the overall effect.  The larger Bondi radius, and the effectively stronger gravity of the core, is the dominant effect that gives faster cooling at lower temperatures.  For different $\beta$ values, $t_{\rm co} \propto F_T^{ \beta + 1/2}$, so the slope of the dust opacity is an important moderating effect on the cooling.

@@

\begin{equation}
\label{eq:delo}
\nabla \di \equiv \frac{3 \kappa(T{\di}) P{\di}}{64 \pi G M \sigma T_{\rm d}^4} L.
\end{equation}
Here $M$ is the total planet mass. Since our analytic model neglects self-gravity, $M=M\co$ and therefore $\nabla\di$ is constant. From equation (\ref{eq:delradan}) and $\delrad=d \ln T / d\ln P$, the temperature profile in the radiative region integrates to
\begin{equation}\label{eq:radTP}
\left(T \over T_{\rm d}\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left( {P \over P_{\rm d}} - 1 \right) \, ,
\end{equation} 
where $\nabla_\infty = 1/(4-\beta)$ is the radiative temperature gradient for $T ,P \rightarrow \infty$.
Applying \Eqs{eq:delradan}{eq:radTP} at the RCB (where $\delrad = \delad$) under the assumption that $P\cb \gg P_{\rm d}$ results in  $T\cb=\chi T\di$ as in \Eq{eq:Tcb}, with $\chi$ defined in \Eq{eq:chi}.

@@

\subsection{Surface Terms}
\label{surfterms}
In this section we check the relevance of the neglected surface terms in \Eq{eq:coolingglobal}.  We first show that accretion energy is only a small correction at the RCB, which is where we apply our cooling model. A rough comparison (ignoring terms of order $\xi$) of  accretion luminosity vs. $\dot{E}$ gives
\begin{equation}
{G M \dot{M} \over R \dot{E}} = {G M  \over R}{dM \over dE} \sim {G M\co^2 \over R_B E}{P\cb \over P_M} \sim \sqrt{R\co \over R_B} \ll 1,
\end{equation} 
where we assume $P\cb \sim P_M$  for a massive atmosphere.  Accretion energy at the protoplanetary surface is thus very weak for marginally self-gravitating atmospheres, and even weaker for lower mass atmospheres.  A similar scaling analysis shows that the work term $P_M \p V_M/\p t$ is similarly weak.  Nevertheless, our numerical calculations include these surface terms in a more realistic and complete model of self-gravitating atmospheres.
